### 2017-02-26 2017mergingmerges/section/A1-Combinators.tex
clean up appendix

Here we show the definitions of some combinators.
We start with simple combinators supported by most streaming systems, and progress to more interesting combinators.
Some combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but can be implemented in our finite stream extension described in~\S\REF.
The fact that segmented versions of these combinators can be implemented is compelling evidence of this.

\subsection{Standard combinators}

\paragraph{Map} is one of the simplest combinators, and is supported almost everywhere.

\begin{code}
map f = process
     ins: as
    outs: bs
    heap: {a}
   label: p0
  blocks: p0 = pull as    a  p1
          p1 = push bs (f a) p2
          p2 = drop as       p0
\end{code}

\paragraph{Filter} is another ubiquitous combinator.

\begin{code}
filter f = process
     ins: as
    outs: bs
    heap: {a}
   label: p0
  blocks: p0 = pull as    a  p1
          p1 = case    (f a) p2 p3
          p2 = push bs    a  p2
          p3 = drop as       p0
\end{code}

\paragraph{Partition} is an interesting combinator since it produces two output streams at once.
The two output streams have different rates.
This sort of splitting is unsupported in pull fusion systems.

\begin{code}
partition f = process
     ins: as
    outs: ts fs
    heap: {a}
   label: p0
  blocks: p0 = pull as    a  p1
          p1 = case    (f a) pT pF
          pT = push ts    a  p3
          pF = push fs    a  p3
          p3 = drop as       p0
\end{code}

\paragraph{Zip} takes two streams and combines them into one.
It is fairly ubiquitous in pull fusion systems, but cannot be supported by push systems.

\begin{code}
zip = process
     ins: as bs
    outs: abs
    heap: {a, b}
   label: p0
  blocks: p0 = pull as    a   p1
          p1 = pull bs    b   p2
          p2 = push abs (a,b) p3
          p3 = drop as        p4
          p4 = drop bs        p0
\end{code}

\paragraph{Scan} performs a prefix sum over an input stream, much like a fold that returns every intermediate value.
For simplicity we are performing just prefix sums, rather than arbitrary scans, but this is easy to generalise.
Here we use two values in the heap; @a@ stores the last seen input and @s@ stores the current sum.
We initialise @s@ to 0 at the start.
Note that this prefix scan includes the seed value 0 in the output; updating the current sum to include the last pulled value happens after the push.

\begin{code}
scan = process
     ins: as
    outs: bs
    heap: {a, s = 0}
   label: p0
  blocks: p0 = pull as a  p1
          p1 = push bs s  p2[s = s + a]
          p2 = drop as    p0
\end{code}

\paragraph{Unfold} creates a new stream from a seed and an iteration function.
The seed has type @s@, the iteration has type @s -> (a,s)@ and the output stream contains @a@s.
This process has only one state which pushes the output, calls the iteration function, and updates the state.
The state/heap is initialised to the first result of the iteration function.

\begin{code}
unfold z f = process
     ins: 
    outs: as
    heap: {s = f z}
   label: p0
  blocks: p0 = push as (fst s) p0[s = f (snd s)]
\end{code}

\paragraph{Merge} takes two sorted inputs and produces a sorted input containing both inputs.
It is the main worker in merge sort.
Merge is a particularly interesting combinator since the access patterns are data dependent; that is, which order it pulls from inputs depends on the values of the inputs.
This is the infinite version of merge; the finite version would have a separate `fixup' stage where after one input has finished, it copies the rest of the other input.

\begin{code}
merge = process
     ins: as bs
    outs: abs
    heap: {a, b}
   label: pI0
  blocks: pI0 = pull as a  pI1
          pI1 = pull bs b  pC0
          pC0 = case (a < b) pA0 pB0

          pA0 = push abs a pA1
          pA1 = drop as    pA2
          pA2 = pull as  a pC0

          pB0 = push abs b pB1
          pB1 = drop bs    pB2
          pB2 = pull bs  b pC0
\end{code}

\subsection{Segmented combinators}

\paragraph{Segmented fold}
takes two input streams; one for the segment lengths, one for the values.
The output stream has the same rate as the segment lengths.

\begin{code}
folds k z = process
     ins: lens vals
    outs: outs
    heap: {count, state, val}
   label: init0
  blocks: init0 = pull lens count go0[state = z]
          go0   = case (count > 0) go1 done0
          go1   = pull vals val    go2
          go2   = jump go0[count = count - 1, state = k state val]
          done0 = push outs state init0
\end{code}

\paragraph{Segmented take}
takes two input streams; the segment descriptor lengths, as well as the underlying values.
It outputs another two streams for the output segment descriptor and values.
The number of elements to take is constant across all segments.

\begin{code}
takes n = process
     ins: lens vals
    outs: olens ovals
    heap: {count, takecount, ix, val}
   label: init0
  blocks: init0 = pull lens count lens0
          lens0 = jump lens1
                [ ix = 0
                , takecount = min count n ]
          lens1 = push olens takecount take0
          take0 = case (ix < takecount) take1 drop0
          take1 = pull vals val take2
          take2 = push ovals val take0[ix = ix + 1]
          drop0 = case (ix < count) drop1 init0
          drop1 = pull vals val drop0[ix = ix + 1]
\end{code}

\paragraph{Segmented append} is very similar to merge.

\begin{code}
apps = process
     ins: Alens Avals Blens Bvals
    outs: olens ovals
    heap: {Alen, Blen, val}
   label: init0
  blocks: init0 = pull Alens Alen init1
          init1 = pull Blens Blen init2
          init2 = push olens  (Alen + Blen) copyA0
          copyA0 = case (Alen > 0) copyA1 copyB0
          copyA1 = pull Avals val copyA2
          copyA2 = push ovals val copyA0[Alen = Alen - 1]
          copyB0 = case (Blen > 0) copyB1 init0
          copyB1 = pull Bvals val copyB2
          copyB2 = push ovals val copyB0[Blen = Blen - 1]
\end{code}

>>>>

Here we show the definitions of some combinators.
We start with simple combinators supported by most streaming systems, and progress to more interesting combinators.
Some standard combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but could be implemented with the finite stream extension described in~\S\ref{s:Finite}.
The fact that segmented versions of these combinators can be implemented is compelling evidence of this.

Many of these combinators take a ``default'' argument, which is used to initialise the heap, but the stored value is never actually read. Ideally these could be left unspecified, or the heap left uninitialised in cases where it is never read.

The map combinator applies a function to every element of the stream.

\begin{alltt}
    map 
        = \(\lambda\) (f : \(\alpha \to \beta\)) (default : \(\alpha\)) (sIn: Stream \(\alpha\)) (sOut: Stream \(\beta\)). 
          \(\nu\) (a: \(\alpha\)) (L0..L2: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sIn }
         , outs:   { sOut }
         , heap:   { a = default }
         , label:  L0
         , instrs: { L0 = pull sIn     a  L1 []
                   , L1 = push sOut (f a) L2 []
                   , L2 = drop sIn        L0 [] } }
\end{code}

Filter returns a new stream containing only the elements that satisfy some predicate.

\begin{alltt}
    filter 
        = \(\lambda\) (f : \(\alpha \to\) Bool) (default : \(\alpha\)) (sIn: Stream \(\alpha\)) (sOut: Stream \(\alpha\)). 
          \(\nu\) (a: \(\alpha\)) (L0..L3: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sIn }
         , outs:   { sOut }
         , heap:   { a = default }
         , label:  L0
         , instrs: { L0 = pull sIn  a     L1 []
                   , L1 = case   (f a)    L2 []  L3 []
                   , L2 = push sOut a     L3 []
                   , L3 = drop sIn        L0 [] } }
\end{code}

\newpage

Partition is similar to filter, but has two output streams: those that satisfy the predicate, and those that do not. Partition is an inherently push-based operation, and cannot be supported by pull streams without buffering.

\begin{alltt}
    partition 
        = \(\lambda\) (f : \(\alpha \to\) Bool) (default : \(\alpha\))
            (sIn: Stream \(\alpha\)) (sOut1: Stream \(\alpha\)) (sOut2: Stream \(\alpha\)). 
          \(\nu\) (a: \(\alpha\)) (L0..L4: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sIn }
         , outs:   { sOut1, sOut2 }
         , heap:   { a = default }
         , label:  L0
         , instrs: { L0 = pull sIn   a    L1 []
                   , L1 = case    (f a)   L2 []  L3 []
                   , L2 = push sOut1 a    L4 []
                   , L3 = push sOut2 a    L4 []
                   , L4 = drop sIn        L0 [] } }
\end{code}

Zip, or zip-with, pairwise combines two input streams.
Zipping is an inherently pull-based operation.

\begin{alltt}
    zipWith 
        = \(\lambda\) (f : \(\alpha \to \beta \to \gamma\)) (default1 : \(\alpha\)) (default2 : \(\beta\))
            (sIn1: Stream \(\alpha\)) (sIn2: Stream \(\beta\)) (sOut: Stream \(\gamma\)). 
          \(\nu\) (a: \(\alpha\)) (b : \(\beta\)) (L0..L4: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sIn1, sIn2 }
         , outs:   { sOut }
         , heap:   { a = default1, b = default2 }
         , label:  L0
         , instrs: { L0 = pull sIn1 a       L1 []
                   , L1 = pull sIn2 b       L2 []
                   , L2 = push sOut (f a b) L3 []
                   , L3 = drop sIn1         L4 []
                   , L4 = drop sIn2         L0 [] } }
\end{code}

Scan is similar to a fold, but instead of returning a single value at the end, it returns an intermediate value for each element of the stream.

\begin{alltt}
    scan 
        = \(\lambda\) (k : \(\alpha \to \beta \to \beta\)) (z : \(\beta\)) (default : \(\alpha\)) (sIn: Stream \(\alpha\)) (sOut: Stream \(\beta\)).
          \(\nu\) (a: \(\alpha\)) (s : \(\beta\)) (L0..L2: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sIn  }
         , outs:   { sOut }
         , heap:   { a = default, s = z }
         , label:  L0
         , instrs: { L0 = pull sIn  a       L1 []
                   , L1 = push sOut s       L2 [ s = f a s ]
                   , L2 = drop sIn          L0 [] } }
\end{code}

\newpage

Segmented fold performs a fold over each nested stream, using a segmented representation.
Here we are representing nested streams using one stream for the lengths of each substream, and another stream containing the values.
The output stream has the same rate as the lengths stream.
It reads a count (@c@) from the lengths stream, setting the fold state to zero (@z@).
Then it reads count times from the values stream, updating the fold state.
Afterwards, it pushes the final fold state, and continues to read a new count.

\begin{alltt}
    folds 
        = \(\lambda\) (k : \(\alpha \to \beta \to \beta\)) (z : \(\beta\)) (default : \(\alpha\))
            (sLens: Stream Nat) (sVals: Stream \(\alpha\)) (sOut: Stream \(\beta\)).
          \(\nu\) (c : Nat) (a: \(\alpha\)) (s : \(\beta\)) (L0..L5: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sLens, sVals }
         , outs:   { sOut }
         , heap:   { c = 0, a = default, s = z }
         , label:  L0
         , instrs: { L0 = pull sLens c      L1 [ s = z ]
                   , L1 = case (c > 0)      L2 []  L4 []
                   , L2 = pull sVals a      L3 []
                   , L3 = drop sVals        L1 [ c = c - 1, s = k s a ]
                   , L4 = push sOut  s      L5 []
                   , L5 = drop sLens        L0 [] } }
\end{code}

Segmented take computes an @n@-length prefix of each nested stream.
It starts by reading a count from the lengths stream, then copies at most @n@ elements.
If there are leftovers, it pulls and discards them, then pulls the next length.

\begin{alltt}
    takes 
        = \(\lambda\) (n : Nat) (default : \(\alpha\))
            (sLens: Stream Nat) (sVals: Stream \(\alpha\))
            (oLens: Stream Nat) (oVals: Stream \(\alpha\)).
          \(\nu\) (c : Nat) (take : Nat) (ix : Nat) (a: \(\alpha\)) (L0..L9: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { sLens, sVals }
         , outs:   { oLens, oVals }
         , heap:   { c = 0, take = 0, ix = 0, a = default }
         , label:  L0
         , instrs: { L0 = pull sLens c      L1 [ ix = 0, take = min count n ]
                   , L1 = push oLens take   L2 []

                   , L2 = case (ix < take)  L3 []  L6 []
                   , L3 = pull sVals a      L4 []
                   , L4 = push oVals a      L5 []
                   , L5 = drop sVals        L2 [ ix = ix + 1 ]

                   , L6 = case (ix < c)     L7 []  L9 []
                   , L7 = pull sVals a      L8 []
                   , L8 = drop sVals        L6 [ ix = ix + 1 ]

                   , L9 = drop sLens        L0 [] } }
\end{code}

\newpage

Segmented append takes two segmented streams as input, and appends each nested stream.
It starts by reading a length from both lengths streams into @a@ and @b@, and pushes the sum of both lengths.
It then copies over @a@ elements from the first values stream, then copies over @b@ elements from the second values stream.

\begin{alltt}
    appends 
        = \(\lambda\) (default : \(\alpha\))
            (aLens: Stream Nat) (aVals: Stream \(\alpha\)) (bLens: Stream Nat) (bVals: Stream \(\alpha\))
            (oLens: Stream Nat) (oVals: Stream \(\alpha\)).
          \(\nu\) (a : Nat) (b : Nat) (v: \(\alpha\)) (L0..L12: Label).
\end{alltt}
\begin{code}
         process
         { ins:    { aLens, aVals, bLens, bVals }
         , outs:   { oLens, oVals }
         , heap:   { a = 0, b = 0, v = default }
         , label:  L0
         , instrs: { L0  = pull aLens  a     L1 []
                   , L1  = pull bLens  b     L2 []
                   , L2  = push oLens (a+b)  L3 []

                   , L3  = case (a > 0)      L4 []  L7 []
                   , L4  = pull aVals v      L5 []
                   , L5  = push oVals v      L6 []
                   , L6  = drop aVals        L3 [ a = a - 1 ]

                   , L7  = case (b > 0)      L8 []  L11[]
                   , L8  = pull bVals v      L9 []
                   , L9  = push oVals v      L10[]
                   , L10 = drop bVals        L7 [ b = b - 1 ]

                   , L11 = drop aLens        L12[]
                   , L12 = drop bLens        L0 [] } }
\end{code}
